{\bf Topology, Geometry and Morphology of the Dark Matter Web}

### 2018-07-25 Draft/Chapter6/concl.tex
final changes

% \begin{itemize}
\subsection{Hierarchical structure of the dark matter web}
% Hierarchical structure of the dark matter web: 
The first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations in the $\Lambda$CDM cosmology reported in Chapter \ref{chapter3} (first shown in \citealt{Ramachandra2015}) shows that at the resolution of the simulation i.e. without additional smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly at the filament crossings, and each filament is embedded in the wall like fabric of the web at the wall crossings.

@@

On the other hand, such definitions for walls, filaments and haloes are empirical, with or without the multistream picture. Haloes, for instance, are often detected using proxies for dark matter particle proximity in position space. These definitions can be simple mass density thresholds, or based on geometry of the dark matter structure (spheroidal, tubular or flat) or topology (connected/isolated).

Our algorithm in \cite{Ramachandra2015} was motivated by the complex and diverse morphologies exhibited by the multi-stream field around a halo. We isolated various halo environments, and checked the variation of areal fraction of multistreams on concentric shells around the halo centers, with radii $r = 0.1 h^{-1} Mpc$ to $5 h^{-1} Mpc$ for each of these haloes. If this fraction scales as $f_{str}(r) \propto r^{-1}$ with respect to the radius $r$, the corresponding multistream region is geometrically identical to that of a wall or a pancake. Such scaling is seen for $n_{str} \cong 3$ regions. This result is completely in agreement of ZA, where the pancakes are the first collapsed regions formed. Our morphology based algorithm also identifies regions with three streams as pancakes.

@@

% Within the dark matter web (regions with $n_{str} > 1$), the walls are regions where $n_{str} \cong 3$. 
Qualitatively, the results are in good agreement with the hierarchical picture of the large scale structures: The multistream web represents a nested structure, consisting of layers with increasing number of streams. The number of streams in the neighboring filaments is higher than in the neighboring walls. However, some of the quantitative results tabulated above (and mentioned in Chapter \ref{chapter3} in detail) may vary depending on the simulation parameters and cosmological models. With the exception of void definition as regions with $n_{str} = 1$, one should be cautious about using the same multistream threshold for walls, filaments and haloes. Instead, the heuristic multistream cut-offs should be verfied or re-calculated for the specific cosmological simulation.

% number of streams is equal to three or a few.
\subsection{Multistream view of the dark matter haloes}

@@

Haloes are perhaps the most investigated dark matter structures in the Universe. Currently over 30 halo finders are found in the literature, with varying levels of agreement within each other. This is not too astounding because apart from the qualitative understanding of haloes being dense, compact structures with large number of gravitational collapses, we do not have a concensus on the definition of a halo. This has resulted in multiple definitions, computational algorithms and post-processing schemes for detecting haloes in simulations.

In our simulation reported in Chapter \ref{chapter3}, simple virial density criterion corresponds to a global threshold of $n_{str} \geq 90$ for DM haloes. This shows a reasonably good correspondence with several state-of-the-art structure finders. Alternatively, a more sophisticated halo finding formulation is done in Chapter \ref{chapter5} assuming that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field.

% \hl{The definition of}

In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites. This halo finding technique is free of any heuristic parameters, and may be extended to delineate filaments, walls, voids, as well as halo sub-structures. Agreement of halo summary statistics (like the halo mass function) of the multistream haloes with traditional halo finders (Friends-of-Friends and Spherical Over-density based techniques) is shown \cite{Ramachandra2017b}. As opposed to several other halo finders, this geometrical approach accounts for non-spherical boundaries, does not find any halo candidates in voids, and minimizes the need for post-processing in the production of halo catalogs for dark matter simulations.

@@

% Topological transitions in the cosmic web: 
Topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields, as shown in Chapter \ref{chapter4}. Single streaming voids occupy over 90 per cent of the dark matter Universe. A single void structure not only percolates the multistream field in all the directions, but also spans over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than simulation resolution reveal tiny pockets of voids that are isolated by membranes of the structure.

Rapid topological transitions are seen in matter density fields as well as multistream fields, i.e., at a given threshold (matter density or multistream), the filling fraction of the largest structure in the Universe dominates the excursion volume fraction. For the multistreaming excursion sets, the percolating structure is significantly thinner than the filaments in over-density excursion approach. We also demonstrated that the matter density field is more fragmented than the multistream field.

@@

The integer value of Flip-flop in Lagrangian field, $n_{ff}(\mathbf{q})$ is increased by every swap of the Eulerian coordinates of the two neighboring DM particles. This number field is estimated in three dimensions by computing the Jacobian  $J(\mathbf{q}, t) = |\frac{\partial\mathbf{x}}{\partial\mathbf{q}}|$  on each particle at each time step. If the sign of the Jacobian changes, the number of flip-flops for the corresponding particles is increased by one. Details of calculation of flip-flop field and its significance in dynamical history of substructure formation is explained in \citep{Shandarin2016}, and summarized in Section \ref{sec:ZA}. This Lagrangian field $n_{ff}(\mathbf{q})$ for a cosmological simulation with $128^3$ DM particles and side length of $100 h^{-1} Mpc$ is shown in the left panel of \autoref{fig:ffq}. Halo environment analysis similar to one in Chapter \ref{chapter5} may allow us to identify and track the formation of DM haloes directly from the Lagrangian space. Alternatively, this also enables back-tracking the DM particles in halos cores and substructures all the way to the initial configurations in the simulation.

\begin{figure}
\begin{minipage}[t]{.99\linewidth}
 \centering\includegraphics[width=5.7cm]{Chapter2/Plots/ffq.pdf} 
\centering\includegraphics[width=5.cm]{Chapter2/Plots/ff0_slice.pdf} 
 \centering\includegraphics[width=5.cm]{Chapter2/Plots/ff1_slice.pdf} 
\end{minipage}\hfill
\caption{Left panel: Lagrangian slice of the Flip-Flop field $n_{ff}(\mathbf{q})$ in a cosmological simulation with $L = 100 h^{-1} Mpc$. The values of $n_{ff}$ varies from $0$ to $15$. Center panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) = 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$. Right panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) > 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$.}
\label{fig:ffq}
\end{figure}

@@

Since the $n_{ff}$ field is defined on particles, mapping the field to Eulerian field is straightforward. Correspondence between the multistream field defined in Eulerian space $n_{str}(\mathbf{x})$ is shown in the middle and right panels of \autoref{fig:ffq}. Most particles that have not undergone a flip-flop are in the single-streaming region. In the multistreaming region, however, consists of particles with $n_{ff} = 0$ and $n_{ff} > 0$. On the other hand, $n_{ff} > 0$ particles are not found in single-streaming voids.

@@

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{Caustic surfaces in a small cosmological simulation of $L=20$ Mpc, $N_p = 32^3$. Caustics are filtered by longest side length $l_{max}$ of the triangle consitituting the caustic surface. Top left: all values of $l_{max}$ (all the caustic surfaces), top right: $ l_{max} \leq 1.8  h^{-1} Mpc$, bottom left: $ l_{max} \leq 1.25  h^{-1} Mpc$, bottom right: $ l_{max} \leq 1.0  h^{-1} Mpc$. Multiple colors in bottom right panel indicate isolated inner caustic surfaces, while the other three panels display connected outer cautic structures.}
\label{fig:caustic_lmax}
\end{figure*}

>>>>

% \begin{itemize}
\subsection{Morphological properties of the dark matter web}
% Hierarchical structure of the dark matter web: 
The first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations in the $\Lambda$CDM cosmology reported in Chapter \ref{chapter3} (first shown in \citealt{Ramachandra2015}) shows that at the resolution of the simulation i.e. without additional smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly at the filament crossings, and each filament is embedded in the wall like fabric of the web at the wall crossings.

@@

On the other hand, such definitions for walls, filaments and haloes are empirical -- with or without the multistream picture. These definitions can be simple mass density thresholds, or based on geometry of the dark matter structure (spheroidal, tubular or flat) or topology (connected/isolated). Our algorithm in \cite{Ramachandra2015} was motivated by the complex and diverse morphologies exhibited by the multi-stream field around a halo. We isolated various halo environments, and checked the variation of areal fraction of multistreams on concentric shells around the halo centers, with radii $r = 0.1 h^{-1} Mpc$ to $5 h^{-1} Mpc$ for each of these haloes. If this fraction scales as $f_{str}(r) \propto r^{-1}$ with respect to the radius $r$, the corresponding multistream region is geometrically identical to that of a wall or a pancake. Such scaling is seen for $n_{str} \cong 3$ regions. This result is completely in agreement of ZA, where the pancakes are the first collapsed regions formed. Our morphology based algorithm also identifies regions with three streams as pancakes.

@@

% Within the dark matter web (regions with $n_{str} > 1$), the walls are regions where $n_{str} \cong 3$. 
Qualitatively, the results are in good agreement with the hierarchical picture of the large scale structures: The multistream web represents a nested structure, consisting of layers with increasing number of streams. The number of streams in the neighboring filaments is higher than in the neighboring walls. However, some of the quantitative results tabulated above (and mentioned in Chapter \ref{chapter3} in detail) may vary depending on the simulation parameters and cosmological models. With the exception of void definition as regions with $n_{str} = 1$, one should be cautious about using the same multistream threshold for walls, filaments and haloes. Instead, the heuristic multistream cut-offs should be verified or re-calculated for the specific cosmological simulation.

% number of streams is equal to three or a few.
\subsection{Geometric view of the multistream haloes}

@@

Haloes are perhaps the most investigated dark matter structures in the Universe. Currently there are over 30 halo finders, with substantial disagreements within each other. This is not too astounding because apart from the qualitative understanding of haloes being dense, compact structures with large number of gravitational collapses, there is no consensus on the definition of a halo. Multiple proposed definitions, computational algorithms and post-processing schemes for detecting haloes in simulations.

The local maxima of multistream fields are the locations of halo centers in the phase space picture. In our simulation reported in Chapter \ref{chapter3}, simple virial density criterion corresponds to a global threshold of $n_{str} \geq 90$ for DM haloes. This shows a reasonably good correspondence with several state-of-the-art structure finders. Alternatively, a more sophisticated halo finding formulation is done in Chapter \ref{chapter5} assuming that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field.

As opposed to several other halo finders, this geometrical approach accounts for non-spherical boundaries and minimizes the need for post-processing in the production of halo catalogs for dark matter simulations. Since the haloes are found only in the $n_{str} > 1$ regions, it is ensured that none of the halo particles are in the void. It is also guaranteed that the halo regions have foldings in the Lagrangian sub-manifold in more than one direction. Finally, this halo finding technique is free of any heuristic parameters, which are often calibrated in other halo finders.

% \hl{The definition of} 
Such methods based on geometry may also be extended to delineate filaments, walls, voids, as well as halo sub-structures. In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites. Agreement of halo summary statistics (like the halo mass function) of the multistream haloes with traditional halo finders (Friends-of-Friends and Spherical Over-density based techniques) is shown \cite{Ramachandra2017b}.

@@

% Topological transitions in the cosmic web: 
Percolation based analyses have been done in cosmological analyses, where one studies the properties of over-dense and under-dense excursion sets. The matter density fields exhibit rapid topological transitions at certain critical density threshold. Similar effect is seen in multistream fields as well, i.e., at a given multistream threshold ($n_{str} \cong 17$ in our cosmological simulations), the filling fraction of the largest structure in the Universe dominates the excursion volume fraction. At this threshold, the excursion set regions transition from isolated patches to connected filamentary structures.

However, the topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields. For the multistreaming excursion sets, the percolating structure is significantly thinner than the percolating filaments in over-density excursion approach. We also demonstrate in Chapter \ref{chapter4} that the matter density field is more fragmented than the multistream field.

While varying density threshold for the under-dense phase of excursion set regions may be used for detecting voids in a mass density field, the void definition in the multistream portrait of the cosmic web is unambiguous. The single-streaming voids occupy over 90 per cent of the dark matter Universe. A single void structure not only percolates the multistream field in all the directions, but also spans over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than the simulation resolution also reveal tiny pockets of voids that are isolated by membranes of the structure.

@@

The integer value of Flip-flop in Lagrangian field, $n_{ff}(\mathbf{q})$ is increased by every swap of the Eulerian coordinates of the two neighboring DM particles. This number field is estimated in three dimensions by computing the Jacobian  $J(\mathbf{q}, t) = |\frac{\partial\mathbf{x}}{\partial\mathbf{q}}|$  on each particle at each time step. If the sign of the Jacobian changes, the number of flip-flops for the corresponding particles is increased by one. Details of calculation of flip-flop field and its significance in dynamical history of substructure formation is explained in \citep{Shandarin2016}, and summarized in Section \ref{sec:ZA}. This Lagrangian field $n_{ff}(\mathbf{q})$ for a cosmological simulation with $128^3$ DM particles and side length of $100 h^{-1} Mpc$ is shown in the bottom panel of \autoref{fig:ffq}. Halo environment analysis similar to one in Chapter \ref{chapter5} may allow us to identify and track the formation of DM haloes directly from the Lagrangian space. Alternatively, this also enables back-tracking the DM particles in halos cores and substructures all the way to the initial configurations in the simulation.

\begin{figure}
\begin{minipage}[t]{.99\linewidth}
\centering\includegraphics[width=7.5cm]{Chapter2/Plots/ff0_slice.pdf} 
\centering\includegraphics[width=7.5cm]{Chapter2/Plots/ff1_slice.pdf} 
 \centering\includegraphics[width=8.0cm]{Chapter2/Plots/ffq.pdf} 
\end{minipage}\hfill
\caption{Top left panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) = 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$. Top right panel: Eulerian picture, only the particles with $n_{ff}(\mathbf{q}) > 0$. are shown. Red regions show regions with $n_{str}(\mathbf{x}) > 1$. Bottom panel: Lagrangian slice of the Flip-Flop field $n_{ff}(\mathbf{q})$ in a cosmological simulation with $L = 100 h^{-1} Mpc$. The values of $n_{ff}$ varies from $0$ to $15$.}
\label{fig:ffq}
\end{figure}

@@

Since the $n_{ff}$ field is defined on particles, mapping the field to Eulerian field is straightforward. Correspondence between the multistream field defined in Eulerian space $n_{str}(\mathbf{x})$ is shown in the top left and top right panels of \autoref{fig:ffq}. Most particles that have not undergone a flip-flop are in the single-streaming region. In the multistreaming region, however, consists of particles with $n_{ff} = 0$ and $n_{ff} > 0$. On the other hand, $n_{ff} > 0$ particles are not found in single-streaming voids.

@@

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{Caustic surfaces in a small cosmological simulation of $L=20$ Mpc, $N_p = 32^3$. Caustics are filtered by longest side length $l_{max}$ of the triangle constituting the caustic surface. Top left: all values of $l_{max}$ (all the caustic surfaces), top right: $ l_{max} \leq 1.8  h^{-1} Mpc$, bottom left: $ l_{max} \leq 1.25  h^{-1} Mpc$, bottom right: $ l_{max} \leq 1.0  h^{-1} Mpc$. Multiple colors in bottom right panel indicate isolated inner caustic surfaces, while the other three panels display connected outer caustic structures.}
\label{fig:caustic_lmax}
\end{figure*}
